---
title: "Community analysis in R using 'vegan' - cheat sheet"
author: "An Bui"
date: "06 November 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction and main takeaways

Today, we'll be working through real data on bird communities in three different habitat types. The data are published in a _really cool_ paper: Adams, Bryce T. and Stephen N. Matthews. "Diverse temperate forest bird assemblages demonstrate closer correspondence to plant species composition than vegetation structure.” _Ecography_. https://doi.org/10.1111/ecog.04487

At the end of this workshop, you should be able to answer the following questions about these data, and apply them to your own data:  

* How speciose are my communities? `specnumber()`  
* How diverse are my communities? `diversity()`  
* How different are my communities in species composition? `adonis()`, `metaMDS()`  
* How do environmental variables predict differences in species composition? `envfit()`
* What variation in species abundance can be explained by environmental variables? `rda()`  
* How is community structure related to specific environmental variables? `cca()`

Let's start by loading in our libraries and data.
```{r set up, message = FALSE}
# libraries
library(tidyverse)
library(here)
library(vegan)
library(ggvegan)

# data
# bird communities
birds <- read_csv(here::here("data", "bird-comm.csv")) %>% 
  column_to_rownames("site")
# environmental variables
env <- read_csv(here::here("data", "env-var.csv"))
# set up a "metadata" frame - will be useful for plotting later!
site_type <- env %>% 
  select(site, landtype)
```

#### What's all this data?

Bird communities were sampled using point counts at 120 points distributed amongst 3 ecological landtypes: 1. dry (ridgetops), 2. riparian (bottomlands), and 3. mixed (hillslopes). The landtypes can be distinguished using their most common tree species, though the site is speciose (~75 tree species).

The `birds` data frame is in the format that data generally have to be in to start doing any analysis, which is to say that samples (or sites) are in rows, and species are columns. **The only entries in the data frame are counts.**  
If you'd like to explore what species are in this set, you can translate alpha codes to common names using [this chart](https://www.birdpop.org/docs/misc/Alpha_codes_eng.pdf).

```{r birds}
head(birds, 3)
```

The `env` data frame contains all the "environmental" variables that were measured, which in this case was tree community composition and canopy structure. An explanation of the columns is below:

Column | Meaning |
------ | ------- |
point  | point count (1-120) |
ELT    | ecological landtype (dry, riparian, mixed) |
stems_ha | total stem density (# of stems per hectare) of both under and overstory stems |
big_stem_bas | large stem basal area (m2/hectare) of trees [stems/trees >/=8 cm diameter at breast height (DBH)] |
canopy_cover | percent canopy cover |
canooy_height | mean canopy height (m) |

### How speciose are my communities? 

`specnumber()` will tell you the number of species within each sample. You can then run an analysis of variance to ask if mean species richness is significantly different across sites.

```{r species richness}
sppr <- specnumber(birds)

aovbirds <- aov(sppr ~ landtype, data = site_type)
summary(aovbirds)
```

Bonus plot with colors painstakingly selected:
```{r}
sppr_df <- sppr %>% 
  enframe() %>% 
  full_join(site_type, by = c("name" = "site"))

pal <- c("lightsalmon1", "gold1", "palegreen4")

sppr_box <- ggplot(sppr_df, aes(x = landtype, y = value, fill = landtype)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) +
  scale_x_discrete(labels = c("dry \n (n = 96)", "mix \n (n = 59)", "riparian \n (n = 55)")) +
  theme(legend.position = "none",
        plot.background = element_rect("white"),
        panel.background = element_rect("white"),
        panel.grid = element_line("grey90"),
        axis.line = element_line("gray25"),
        axis.text = element_text(size = 12, color = "gray25"),
        axis.title = element_text(color = "gray25"),
        legend.text = element_text(size = 12)) + 
  labs(x = "Ecological landtype",
       y = "Number of species",
       title = "Species richness")
sppr_box
```

### How diverse are my communities? `diversity()`

There are _so very many_ diversity metrics out there, but I've only included Shannon diversity here, takes into account species abundance and evenness: $$ H = -\sum_{i=1}^{R}p_iln(p_i) $$

where $R$ is total richness, $p_i$ is the proportion of $R$ of the $i$th species. The `diversity()` function in `vegan` will calculate Shannon, Simpson, and Fisher's alpha - just select which one you want in the arguments. Here, I'll calculate Shannon diversity for each site, then plot mean Shannon diversity per landtype.

```{r shannon diversity, message = FALSE}
shannondiv <- birds %>% 
  # calculate Shannon diversity for each sample - Shannon is the default
  diversity() %>% 
  # put all those calculations into a tibble
  enframe() %>% 
  # rename columns for ease of joining
  rename(site = name,
         shan_div = value) %>% 
  # join with site_type
  full_join(site_type, ., by = "site") %>% 
  # group by landtype
  group_by(landtype) %>% 
  # calculate mean and standard error of diversity
  summarize(mean = mean(shan_div),
            err = sd(shan_div)/sqrt(length(shan_div)))
shannondiv

shandivplot <- ggplot(shannondiv, aes(x = landtype, y = mean, fill = landtype)) +
  geom_col(color = "black") +
  geom_errorbar(aes(ymin = mean - err, ymax = mean + err))
shandivplot
```

Another bonus plot:
```{r}
nicer_shannondivdf <- shannondiv %>% 
  mutate(mean = round(mean, 2)) %>% 
  dplyr::mutate(label = "mean") %>% 
  unite("mean_label", label, mean, sep = " = ", remove = FALSE)

nicer_shandivplot <- ggplot(nicer_shannondivdf, aes(x = landtype, y = mean, fill = landtype)) +
  geom_col(color = "black") +
  scale_fill_manual(values = pal) +
  geom_errorbar(aes(ymin = mean - err, ymax = mean + err), width = 0.5) +
  geom_text(aes(x = landtype, y = mean + err + 0.1, label = mean_label)) +
  scale_x_discrete(labels = c("dry \n (n = 96)", "mix \n (n = 59)", "riparian \n (n = 55)")) +
  scale_y_continuous(limits = c(0, 2.75), expand = c(0,0)) +
  theme(legend.position = "none",
        plot.background = element_rect("white"),
        panel.background = element_rect("white"),
        panel.grid = element_line("grey90"),
        axis.line = element_line("gray25"),
        axis.text = element_text(size = 12, color = "gray25"),
        axis.title = element_text(color = "gray25"),
        legend.text = element_text(size = 12)) + 
  labs(x = "Ecological landtype",
       y = "Mean Shannon diversity",
       title = "Shannon diversity")

nicer_shandivplot
```


### How different are my communities in species composition?

Assessing differences in community composition is done with **per**mutational **M**ultivariate **An**alysis **o**f **Va**riance, or perMANOVA. These tests are done on distances, meaning that they assess the differences between communities based on dissimilarity. With perMANOVA, the null hypothesis is that the centroids of your groups (in ordination space as defined by the dissimilarity measure you've chosen) are equivalent for all groups. In other words, you are asking if, following some measure of (dis)similarity, the community composition of sites between groups is the same.  

The function `adonis()` will do this for you very easily.

```{r permanova}
bird_perm <- adonis(birds ~ landtype, data = env)
# HELP how to interpret F and R2
bird_perm
```

According to the perMANOVA, the community composition of sites between groups is significantly different. But what does this actually look like?

#### What are the effects of environmental variation on community composition?

Ordination is a method for describing the effects of environmental variation (generally between sites) on community composition. Like diversity, there are _very many_ different methods of ordination. I'm still wrapping my head around how to pick the best method for any given dataset, but the way most people tend to choose a method is to go with whatever has been done in their field. 

##### Non-metric Multidimensional Scaling (NMDS)

Most community ecologists use **N**on-metric **M**ultidimensional **S**caling (NMDS), so we'll start there first. You can imagine NMDS as a reduction of axes, where all your "axes" are the species within a sample, and each sample exists relative to others on the axes. For this dataset, imagine an axis that describes relative abundance of ACFL (Acadian Flycatchers) - all points exist somewhere on that axis relative to the others. Now consider another axis describing the relative abundance of KEWA (Kentucky Warblers) - all points still exist somewhere on that axis, but now there are two axes describing the position of your points. **NMDS** allows you to collapse all these species axes (in this case, 48) into 2 to plot in cartesian space in order to visualize the differences between samples and sites. We'll do this by using `metaMDS()`. 

```{r NMDS}
# do the NMDS - again, play around with arguments based on your own needs!
bird_NMDS <- metaMDS(birds)

# look at the results
bird_NMDS

# points: sample scores
# dims: number of NMS axes
# stress: stress of final solutions
# data: what was ordinated and any transformations
# distance: distance metric used
# converged: whether or not solution converged
# tries: number of configurations tried
# species: scores of variables
# call: how function was called

# look at the stress plot (HELP how do you interpret this)
stressplot(bird_NMDS)

plot(bird_NMDS)
# circles are species, crosses are sites
```

```{r}
# 1. get a ggplot ready data frame
# extract the points from the results of metaMDS()
plot_df <- as_tibble(bird_NMDS$points) %>% 
  # bind the results with your metadata frame
  bind_cols(site_type, .)

# 2. plot in ggplot!
NMDSplot <- ggplot(plot_df, aes(x = MDS1, y = MDS2, color = landtype)) +
  geom_point() +
  stat_ellipse()
NMDSplot

# Bonus: a prettier plot
nicer_NMDSplot <- ggplot(plot_df, aes(x = MDS1, y = MDS2, color = landtype, shape = landtype)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = pal) +
  stat_ellipse(linetype = 2, size = 1)
nicer_NMDSplot
```

##### What's standardization?

Sometimes your ordination doesn't converge on a single solution - that's ok! There are many reasons why that might happen (e.g. very speciose communities, many rare species). However, you can **standardize** your communities to make the calculation work, meaning that you transform your count data relative to other samples in your dataset.
```{r}
# standardize by total number of counts - summing across species within a site, everything adds up to 1
bird_t <- decostand(birds, method = "total")

bird_t_NMDS <- metaMDS(bird_t)
stressplot(bird_t_NMDS)

t_plot_df <- as_tibble(bird_t_NMDS$points) %>% 
  bind_cols(site_type, .)

nicer_t_NMDSplot <- ggplot(t_plot_df, aes(x = MDS1, y = MDS2, color = landtype, shape = landtype)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = pal) +
  stat_ellipse(linetype = 2, size = 1)
nicer_t_NMDSplot
```

```{r}
envfit_ord <- envfit(bird_NMDS, env = env)
envfit_ord
plot(envfit_ord)

vectors_df <- envfit_ord %>%
  scores(., display = "vectors") %>%
  as.data.frame()

with_vectors <- ggplot(plot_df) +
  geom_point(aes(x = MDS1, y = MDS2, color = landtype, shape = landtype), size = 3, alpha = 0.8) + 
  stat_ellipse(aes(x = MDS1, y = MDS2, color = landtype), linetype = 2, size = 1) +
  scale_color_manual(values = pal) +
  geom_text(data = vectors_df, aes(x = NMDS1, y = NMDS2, label = rownames(vectors_df))) +
  geom_segment(data = vectors_df, aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2)) +
  coord_fixed()

with_vectors
```


#### Principal Components Analysis (PCA)

**P**rincipal **C**omponents **A**nalysis (PCA) is di

```{r}
bird_PCA <- rda(birds)
bird_PCA
summary(bird_PCA)
screeplot(bird_PCA, bstick = TRUE, type = "lines")
autoplot(bird_PCA)
biplot(bird_PCA, scaling = -1)

scores1 <- data.frame(bird_PCA$CA$u) %>% 
  cbind(site_type, .)

scores2 <- scores(bird_PCA, display = "sites")
  
vscores <- data.frame(bird_PCA$CA$v)

ggplot(scores, aes(x = PC1, y = PC2, color = landtype, shape = landtype)) +
  geom_point()

ggplot(vscores) +
  geom_text(aes(x = PC1, y = PC2, label = rownames(vscores))) +
  geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2))

ggplot(data = scores) +
  geom_point(aes(x = PC1, y = PC2, color = landtype, shape = landtype)) +
  scale_color_manual(values = pal) +
  scale_x_continuous(limits = c(-0.2, 0.2)) +
  scale_y_continuous(limits = c(-0.25, 0.25)) +
  geom_text(data = vscores, aes(x = PC1, y = PC2, label = rownames(vscores))) +
  geom_segment(data = vscores, aes(x = 0, y = 0, xend = PC1, yend = PC2))
```

```{r}
RDA <- rda(birds ~ big_stem_bas + canopy_cover, env)
RDA

RDA_coords <- scores(RDA, display = c("sites", "species"))
plot(RDA)
autoplot(RDA)
```

```{r}
birdCCA <- cca(birds ~ big_stem_bas * canopy_cover, data = env)
birdCCA

plot(birdCCA)
ggcca <- autoplot(birdCCA)
ggcca

data <- fortify(birdCCA) %>% 
  filter(Score == "sites") %>% 
  full_join(., site_type, by = c("Label" = "site"))

ggplot(data, aes(x = CCA1, y = CCA2, color = landtype)) +
  geom_point() +
  scale_color_manual(values = c("lightsalmon1", "gold1", "palegreen4")) +
  stat_ellipse()
```


### Bibliography and further reading

- I drew from Jim Baxter's materials for a class entitled [Methods in Ecology, Evolution, and Conservation](http://irt-pw-cp1.irt.csus.edu/jbaxter/courses/methods-in-eec/) at Sacramento State for this excellent [definition of ordination](csus.edu/indiv/b/baxterj/bio%20221b/ordination.pdf).  
- Many of these functions are described in more technical language in [Jari Oksanen's own tutorial](http://cc.oulu.fi/~jarioksa/opetus/metodi/vegantutor.pdf) for using `vegan` for ordination.  
- Michael M. Palmer's [Ordination Methods - an overview](http://ordination.okstate.edu/overview.htm) contains descriptions of _many_ different ordination methods that I revisit often.
- This [comparison of different techniques](http://www.umass.edu/landeco/teaching/multivariate/schedule/summary.handouts.pdf) in multivariate statistics is basically a flowchart for deciding which test is best for your questions and data. It's a lifesaver!
